---
title: "Model Extension"
author: "Kirisanth"
date: "Tuesday, November 18, 2014"
output: html_document
---
**Initial Model**
White tail deer and Plant Model
predator-prey model
Can be applied to model polar bears and seal interaction

$$ P(t + 1) = P(t) + aP(t) - cP(t)D(t) $$
$$ D(t + 1) = D(t) + ecP(t)D(t) - fD(t) - h $$ 
where:\
$a$ is the growth rate of plants in the absence of deer \
$c$ is the per-deer-plant-encounter rate \
$e$ is the efficiency of deer at converting captured plants into resources \
$f$ is the per-deer mortality rate \
$h$ is the permitted hunting rate \


Solve for fixed points \
$P{\star} = P{\star}(1 + a - cD{\star})$ \
$D{\star} = D{\star}(1 + ecP{\star} - f - h)$ \

Fixed Points: \
$(0, 0)$ \
$((-1+f+h)/ec, (1+a)/c)$ \


A fixed point is stable if all of the eigenvalues of this matrix (evaluating at the fixed point) have absolute value less than one.\
The Jacobian for the predator-prey model is \

$\begin{equation*}
  \label{eq:1}
  \begin{pmatrix}
    \frac{\partial D(t+1)}{\partial D(t)} & 
    \frac{\partial D(t+1)}{\partial P(t)} \\
    \frac{\partial P(t+1)}{\partial D(t)} & 
    \frac{\partial P(t+1)}{\partial P(t)} \\
  \end{pmatrix}
\end{equation*}$

Jacobian for above equations\

$\begin{equation*}
  \label{eq:2}
  \begin{pmatrix}
    1 + a - cD &
    -cP \\
    ecD &
    ecP - f \\
  \end{pmatrix}
\end{equation*}$

For $(0, 0)$ \
$\begin{equation*}
  \label{eq:3}
  \begin{pmatrix}
    1 + a &
    0 \\
    0 &
    - f \\
  \end{pmatrix}
\end{equation*}$

For $((-1+f+h)/ec, (1+a)/c)$ \
$\begin{equation*}
  \label{eq:4}
  \begin{pmatrix}
    1 + a - c((1+a)/c) &
    -c((-1+f+h)/ec) \\
    ec((1+a)/c) &
    ec((-1+f+h)/ec) - f \\
  \end{pmatrix}
\end{equation*}$

$\begin{equation*}
  \label{eq:5}
  \begin{pmatrix}
    0 &
    (1-f-h)/e \\
    e(1+a) &
    (h-1) \\
  \end{pmatrix}
\end{equation*}$

From wolfram alpha: \
eigenvalues \
$\lambda_{1} = 1/2(-\sqrt{-4af - 4ah + 4a - 4f + h^2 - 6h + 5} + h - 1)$ \
$\lambda_{2} = 1/2(\sqrt{-4af - 4ah + 4a - 4f + h^2 - 6h + 5} + h - 1)$ \

stable if: \
$\lambda_{1} < 1$ \
$\lambda_{2} < 1$ \


```{r}
P <- 1:50 

P[1] = 100

D <- 1:50

D[1] = 1

#Plant growth rate without deer.
a <- 2
#Plant deer encounter rate
c <- 0.5
#efficiency for deer to convert plant to deer
e <- 0.3
#mortality rate deer
f <- 0.6
#hunting rate deer
h = 0

for (i in 1:49) {
  P[i+1] <- P[i] + a*P[i] - c*P[i]*D[i]
  D[i+1] <- D[i] + e*c*P[i]*D[i] - f*D[i] - h

}

print(P)
print(D)

```

Plots:

```{r, echo=FALSE}
plot(D, ylim=c(-4.780400e+01,4.780400e+01),type="l")
plot(P)
```

<!--
**Extended Model**
Polar bear Population and Seal hunting
predator-prey model

**Preliminary question**
Polar bears once abundant in the artic circle are now facing extinction. They are found in Canada, Greenland, Russia, and Alaska. Polar bears are an apex predator. One of their primary food source is the seal. If hunting of seals were to increase it could hurt polar bear populations. Here we will model polar bear populations with seal populations. What should the quota of seal hunting be set to, to keep polar bear populations stable.

**Variables** \
$P(t)$ number of polar bears \
$S(t)$ numbner of seals \
$t$ time \


**Extended model** \
Equation: \

Seals
$$ S(t + 1) = S(t) + aS(t) - cS(t)P(t) - h$$
Polar bears
$$ P(t + 1) = P(t) + ecS(t)P(t) - fP(t)$$ 



```{r}
S <- 1:50
S[1] = 100

Pl <- 1:50
Pl[1] = 1

#Seal growth rate without polar bears.
a <- 2
#Seal polar bear encounter rate
c <- 0.5
#efficiency for polar bears to convert seals to another polar polar bear
e <- 0.3
#mortality rate polar bear
mp <- 0.6
#hunting rate seals
h = 0

for (i in 1:50) {
  #seals
  S[i+1] <- S[i] + a*S[i] - c*S[i]*Pl[i] - h
  #polar bears
  Pl[i+1] <- Pl[i] + e*c*S[i]*Pl[i] - f*Pl[i]

}
```

```{r, echo=FALSE}
plot(D)
plot(Pl)
```

-->

**Extended Model**
Polar bear Population and Seal hunting
predator-prey model

**Preliminary question**
Polar bears once abundant in the artic circle are now facing extinction. They are found in Canada, Greenland, Russia, and Alaska. Polar bears are an apex predator. One of their primary food source is the seal. If hunting of seals were to increase it could hurt polar bear populations. Here we will model polar bear populations with seal populations. What should the quota of seal hunting be set to, to keep polar bear populations stable.

**Variables** \
$S(t)$ Seal population \
$P(t)$ Polar bear population \
$H(t)$ Human population\
$C_{S}(t)$ Seal conservation population\
$C_{P}(t)$ Polar bear conservation population\
$I(t)$ Industrial activity and shipping through northwest passage\
$t$ Time \

**Extended model** \
Equation: \

Polar bear population vs seals population and rising artic population

Seal population
$$ S(t + 1) = S(t) + aS(t) - cS(t)P(t) - hS(t)H(t) - i_{s}P(t)I(t)$$
Polar bear population
$$ P(t + 1) = P(t) + ecS(t)P(t) - fP(t) - i_{p}P(t)I(t)$$ 
Human Population
$$ H(t+1) = H(t) + ecS(t)H(t) + ecP(t)H(t) - mH(t) + i_{h}H(t)I(t)$$
Seal Conservation
$$ C_{S}(t+1) = C_{S}(t) + r_{S}C_{S}(t)(1 - C_{S}(t)/K_{SC}) - m_{S}C_{S} $$
Polar bear Conservation
$$ C_{P}(t+1) = C_{P}(t) + r_{P}C_{P}(t)(1 - C_{P}(t)/K_{PC}) - m_{P}C_{P} $$
Industrial Shipping through Northwest Passage
$$ I(t+1) = I(t) + i_{g}I(t)(1-I(t)/EC)$$

//TODO: need to fix above equations

where: \
$s_{g}$ - Seal growth rate \
$c_{ps}$ - Polar bear seal interaction rate \
$c_{hs}$ - Human Seal interaction rate \
$i_{s}$ - Population decline resulting from increasing economic activity \
\
$e_{ps}$ - Polar bear efficiency to convert one seal to another polar bear \
$c_{ps}$ - Polar bear seal interaction rate \
$c_{hp}$ - Polar bear human interaction rate \
$f$$d_{p}$ - Polar bear mortality rate \
$i_{p}$ - Population decline resulting from increasing economic activity \
\
$e_{hs}$ - Human efficiency to convery one seal to a human \
$e_{hp}$ - Human efficiency to convert one polar bear to a human \
$c_{hs}$ - Human seal interaction rate\
$c_{hp}$ - Human polar bear interaction rate\
$d_{h}$ - Human mortality rate \
$i_{h}$ - Population growth resulting from increasing economic activity \
\
$r_{S}$ - Seal reservation growth rate \
$m_S$ - Seal migration out rate \
$K_{SC}$ - Seal reserve capacity \
\
$r_{P}$ - Polar bear reservation growth rate \
$m_P$ - Polar bear migration out rate \
$K_{PC}$ - Polar bear reserve capacity \
\
$i_{g}$ - Industrial growth rate factors include melting of ice in northwest passage area \
$EC$ - Economic capacity represents diminishing returns in GDP growth \

*Assumptions*


```{r}
S <- 1:50
P <- 1:50
H <- 1:50
CS <- 1:50
CP <- 1:50

S[1] = 20
P[1] = 2
H[1] = 2
CS[1] = 2
CP[1] = 2

#Equation 1 seal population
sg <- 1.3 #Seal growth rate without polar bear or humans.
sp <- 0.2 #seal-polar bear encounter rate
sh <- 0.1 #seal-human hunting rate

#Equation 2 polar bear population
eps <- 0.10 #polar bear efficiency to convert one seal to another polar bear
mp <- 0.10 #polar bear mortality rate

#Equation 3 human population
ehs <- 0.05 #human efficiency to convert seals to human
ehp <- 0.1 #human efficiency to convert polar bears to human
sh <- 0.2 #seal human encounter rate
ph <- 0.2 #polar bear human encounter rate
mh <- 0.1 #human mortality rate

#Equation 4
KS <- 40 #Capacity
csg <- 0.8 #seal growth rate in conservation area
ts <- 0.8 #seal migration out rate

#Equation 5
KP <- 20 #Capacity
cpg <- 0.1 #polar growth rate in conservation area
tp <- 0.05 #polar bear migration out rate

for (t in 1:49) {
  
  #Seal population
  S[t + 1] <- S[t] + sg*S[t] - sp*S[t]*P[t] - sh*S[t]*H[t] + ts*CS[t]
  #Polar bear population
  P[t + 1] <- P[t] + eps*sp*S[t]*P[t] - mp*P[t] + tp*CP[t]
  #Human Population
  H[t+1] <- H[t] + ehs*sh*S[t]*H[t] + ehp*ph*P[t]*H[t] - mh*H[t]
  #Seal Conservation
  CS[t+1] <- CS[t] + csg*CS[t]*(1 - CS[t]/KS) - ts*CS[t]  #1 dimensional
  #Polar bear Conservation
  CP[t+1] <- CP[t] + cpg*CP[t]*(1 - CP[t]/KP) - tp*CP[t] #1 dimensional
}


```

Plots:

```{r, echo=FALSE}
t=seq(0,49,1)
plot(t,S,type="l",col="red", ylim=c(-300, 300))
lines(t,P,col="green")
lines(t,H, col="purple")
lines(t,CS, col="brown")
lines(t,CP, col="black")
grid()
legend("bottomleft", legend=c("Polar bear", "Seal","Human","Polar Con", "Seal Con"), lwd=c(3,3), col=c("red","green","purple","brown", "black"))
```

**References**

