---
title: "ModelExtension"
author: "Kirisanth"
date: "Monday, December 01, 2014"
output: html_document
---

NOT BEING USED


```{r}
S <- 1:50 #Seal population
P <- 1:50 #Polar bear population
H <- 1:50 #Human population
CS <- 1:50 #Seal conservation population
CP <- 1:50 #Polar bear conservation population
I <- 1:50 #Industrial shipping through northwest passage

S[1] = 40 #Initial Seal population
P[1] = 5 #Initial polar bear population
H[1] = 5 #Initial human population
CS[1] = 2 #Initial seal conservation population
CP[1] = 2 #Initial polar bear conservation population
I[1] = 1 #Initial amount of Industrial activity (shipping)

#Equation 1 seal population
rg <- 1.0 #Seal growth rate without polar bear or humans.
cps <- 0.1 #seal-polar bear encounter rate
is <- 0.1 #industrial effect on seal population
SC <- 100 #seal carrying capacity

#Equation 2 polar bear population
eps <- 0.2 #polar bear efficiency to convert one seal to another polar bear
dp <- 0.1 #polar bear mortality rate
ip <- 0.1 #industrial effect on polar bear population

#Equation 3 human population
ehs <- 0.01 #human efficiency to convert seals to human
ehp <- 0.01 #human efficiency to convert polar bears to human
chs <- 0.01 #seal human encounter rate
chp <- 0.01 #polar bear human encounter rate
dh <- 0.5 #human mortality rate
ih <- 0.05 #industrial effect on human population

#Equation 4
KSC <- 80 #Capacity
rs <- 1.1 #seal growth rate in conservation area
ms <- 0.9 #seal migration out rate

#Equation 5
KPC <- 30 #Capacity
rp <- 1.1 #polar growth rate in conservation area
mp <- 0.5 #polar bear migration out rate

#Equation 6
ri <- 0.10 #industrial shipping growth rate
EC <- 100 #Economic constraint represents diminishing returns of economic growth

for (t in 1:49) {
  print((1-(S[t]/SC)))
  #Seal population
  S[t + 1] <- S[t] + rg*S[t] - cps*S[t]*P[t] - chs*S[t]*H[t] - is*S[t]*I[t] + ms*CS[t]
  #Polar bear population
  P[t + 1] <- P[t] + eps*cps*S[t]*P[t] - dp*P[t] - chp*P[t]*H[t] - ip*P[t]*I[t] + mp*CP[t]
  #Human Population
  H[t+1] <- H[t] + ehs*chs*S[t]*H[t] + ehp*chp*P[t]*H[t] + ih*H[t]*I[t] - dh*H[t]
  #Seal Conservation
  CS[t+1] <- CS[t] + rs*CS[t]*(1 - CS[t]/KSC) - ms*CS[t]  #1 dimensional
  #Polar bear Conservation
  CP[t+1] <- CP[t] + rp*CP[t]*(1 - CP[t]/KPC) - mp*CP[t] #1 dimensional
  #Industial Shipping through Northwest Passage
  I[t+1] <- I[t] + ri*I[t]*(1 - I[t]/EC)
}

t=seq(0,49,1)
plot(t,S,type="l",col="blue")
# plot(t,S,type="l",col="blue", ylim=c(-20, 45))
# plot(t,S,type="l",col="blue", ylim=c(-20, 60))
lines(t,P,col="red")
lines(t,H, pch=22, lty=2, col="green")
lines(t,CS, pch=22, lty=2, col="blue")
lines(t,CP, pch=22, lty=2, col="red")
lines(t,I, col="green")
grid()
#legend("bottomleft", legend=c("Polar bear", "Seal","Human","Polar Con", "Seal Con"), lwd=c(3,3), col=c("red","green","purple","brown", "black", "yellow"))
```


**Extended Model**
Polar bear population, with Seal hunting, and rising arctic economic activity 
predator-prey model extended to include other factors


**Preliminary question**
Polar bears once abundant in the artic circle are now facing extinction. They are found in Canada, Greenland, Russia, and Alaska. Polar bears are an apex predator. One of their primary food source is the seal. If hunting of seals and polar bears were to increase it could hurt polar bear populations. Along with the hunting as sea ice in the arctic ocean melts, the northwest passage is becoming a more viable option for shipping that require crossing the american continent (Panama Canal).  Here we will model polar bear populations with seal populations with rising economic activity in the arctic. What should the quota of shipping through the arctiv be set to, to keep polar bear and seal populations stable.

**Variables** \
$S(t)$ Seal population \
$P(t)$ Polar bear population \
$H(t)$ Human population\
$C_{S}(t)$ Seal conservation population\
$C_{P}(t)$ Polar bear conservation population\
$I(t)$ Industrial activity and shipping through northwest passage\
$t$ Time \

**Extended model** \
Equation: \

Polar bear population vs seals population and rising artic population

Seal population
$$ S(t + 1) = S(t) + r_{g}S(t) - c_{ps}S(t)P(t) - c_{hs}S(t)H(t) - i_{s}P(t)I(t) + m_{S}C_{S}(t)$$
Polar bear population
$$ P(t + 1) = P(t) + e_{ps}c_{ps}S(t)P(t) - d_{p}P(t) - c_{hp}P(t)H(t) - i_{p}P(t)I(t) + m_{P}C_{P}(t)$$ 
Human Population
$$ H(t+1) = H(t) + e_{hs}c_{hs}S(t)H(t) + e_{hp}c_{hp}P(t)H(t) - d_{h}H(t) + i_{h}H(t)I(t)$$
Seal Conservation
$$ C_{S}(t+1) = C_{S}(t) + r_{S}C_{S}(t)(1 - C_{S}(t)/K_{SC}) - m_{S}C_{S}(t)$$
Polar bear Conservation
$$ C_{P}(t+1) = C_{P}(t) + r_{P}C_{P}(t)(1 - C_{P}(t)/K_{PC}) - m_{P}C_{P}(t)$$
Industrial Shipping through Northwest Passage
$$ I(t+1) = I(t) + r_{i}I(t)(1-I(t)/EC)$$

where: \
$r_{g}$ - Seal growth rate \
$c_{ps}$ - Polar bear seal interaction rate \
$c_{hs}$ - Human Seal interaction rate \
$i_{s}$ - Population decline resulting from increasing economic activity \
\
$e_{ps}$ - Polar bear efficiency to convert one seal to another polar bear \
$c_{ps}$ - Polar bear seal interaction rate \
$c_{hp}$ - Polar bear human interaction rate \
$d_{p}$ - Polar bear natural mortality rate \
$i_{p}$ - Population decline resulting from increasing economic activity \
\
$e_{hs}$ - Human efficiency to convery one seal to a human \
$e_{hp}$ - Human efficiency to convert one polar bear to a human \
$c_{hs}$ - Human seal interaction rate\
$c_{hp}$ - Human polar bear interaction rate\
$d_{h}$ - Human mortality rate \
$i_{h}$ - Population growth resulting from increasing economic activity \
\
$r_{S}$ - Seal reservation growth rate \
$m_S$ - Seal migration out rate \
$K_{SC}$ - Seal reserve capacity \
\
$r_{P}$ - Polar bear reservation growth rate \
$m_P$ - Polar bear migration out rate \
$K_{PC}$ - Polar bear reserve capacity \
\
$r_{i}$ - Industrial growth rate factors include melting of ice in northwest passage area \
$EC$ - Economic capacity represents diminishing returns in GDP growth \

*Assumptions and Explanation of Constants*


```{r}
S <- 1:50 #Seal population
P <- 1:50 #Polar bear population
H <- 1:50 #Human population
CS <- 1:50 #Seal conservation population
CP <- 1:50 #Polar bear conservation population
I <- 1:50 #Industrial shipping through northwest passage

S[1] = 40 #Initial Seal population
P[1] = 10 #Initial polar bear population
H[1] = 10 #Initial human population
CS[1] = 2 #Initial seal conservation population
CP[1] = 2 #Initial polar bear conservation population
I[1] = 1 #Initial amount of Industrial activity (shipping)

#Equation 1 seal population
rg <- 1.0 #Seal growth rate without polar bear or humans.
cps <- 0.1 #seal-polar bear encounter rate
is <- 0.1 #industrial effect on seal population
SC <- 100 #seal carrying capacity

#Equation 2 polar bear population
eps <- 0.2 #polar bear efficiency to convert one seal to another polar bear
dp <- 0.1 #polar bear mortality rate
ip <- 0.1 #industrial effect on polar bear population

#Equation 3 human population
ehs <- 0.01 #human efficiency to convert seals to human
ehp <- 0.01 #human efficiency to convert polar bears to human
chs <- 0.01 #seal human encounter rate
chp <- 0.01 #polar bear human encounter rate
dh <- 0.5 #human mortality rate
ih <- 0.05 #industrial effect on human population

#Equation 4
KSC <- 80 #Capacity
rs <- 1.1 #seal growth rate in conservation area
ms <- 0.9 #seal migration out rate

#Equation 5
KPC <- 30 #Capacity
rp <- 1.1 #polar growth rate in conservation area
mp <- 0.5 #polar bear migration out rate

#Equation 6
ri <- 0.10 #industrial shipping growth rate
EC <- 100 #Economic constraint represents diminishing returns of economic growth

for (t in 1:49) {
  print((1-(S[t]/SC)))
  #Seal population
  S[t + 1] <- S[t] + rg*S[t] - cps*S[t]*P[t] - chs*S[t]*H[t] - is*S[t]*I[t] + ms*CS[t]
  #Polar bear population
  P[t + 1] <- P[t] + eps*cps*S[t]*P[t] - dp*P[t] - chp*P[t]*H[t] - ip*P[t]*I[t] + mp*CP[t]
  #Human Population
  H[t+1] <- H[t] + ehs*chs*S[t]*H[t] + ehp*chp*P[t]*H[t] + ih*H[t]*I[t] - dh*H[t]
  #Seal Conservation
  CS[t+1] <- CS[t] + rs*CS[t]*(1 - CS[t]/KSC) - ms*CS[t]  #1 dimensional
  #Polar bear Conservation
  CP[t+1] <- CP[t] + rp*CP[t]*(1 - CP[t]/KPC) - mp*CP[t] #1 dimensional
  #Industial Shipping through Northwest Passage
  I[t+1] <- I[t] + ri*I[t]*(1 - I[t]/EC)
}

t=seq(0,49,1)
plot(t,S,type="l",col="blue")
# plot(t,S,type="l",col="blue", ylim=c(-20, 45))
# plot(t,S,type="l",col="blue", ylim=c(-200, 200))
lines(t,P,col="red")
lines(t,H, pch=22, lty=2, col="green")
lines(t,CS, pch=22, lty=2, col="blue")
lines(t,CP, pch=22, lty=2, col="red")
lines(t,I, col="green")
grid()
#legend("bottomleft", legend=c("Polar bear", "Seal","Human","Polar Con", "Seal Con"), lwd=c(3,3), col=c("red","green","purple","brown", "black", "yellow"))
```

Plots:

```{r, echo=FALSE}
t=seq(0,49,1)
plot(t,S,type="l",col="blue", ylim=c(-50, 300))
lines(t,P,col="red")
lines(t,H, pch=22, lty=2, col="green")
lines(t,CS, pch=22, lty=2, col="blue")
lines(t,CP, pch=22, lty=2, col="red")
lines(t,I, col="green")
grid()
legend("topleft", legend=c("Polar bear", "Seal","Human","Polar Con", "Seal Con"), lwd=c(3,3), col=c("red","green","purple","brown", "black", "yellow"))
```


**Conclusion**


**References**

http://www.polarbearhabitat.ca/research
http://geology.com/articles/northwest-passage.shtml
http://www.planetwater.ca/research/climpact/impacts/shipping.html
http://www.theglobeandmail.com/report-on-business/breakthrough/will-cold-dark-northwest-passage-see-more-ships/article16231502/
https://www.ocf.berkeley.edu/~reinholz/ed/07fa_m155/lectures/discrete-time_dynamical_systems.pdf
http://www.wired.co.uk/news/archive/2013-03/05/trans-arctic-shipping-routes-opening-warming
http://www.wwf.ca/conservation/arctic/a_new_frontier_for_shipping/
http://www.polarbearsinternational.org/about-polar-bears/essentials/hunting-and-eating
http://www.plosone.org/article/info%3Adoi%2F10.1371%2Fjournal.pone.0041429#s1
http://www.worldwildlife.org/species/polar-bear
